\chapter{MDP Formulation of Gate Assignment}
\label{chap:mdp}

\section{Problem Description}

An airport has $G$ gates available for aircraft turnaround operations and $R$ runways for arrivals.
Flights arrive according to a stochastic schedule over a finite time horizon $T$. Each flight lands on a specific runway and requires assignment to a gate for a duration that depends on both the aircraft type and the taxiing distance between the runway and the assigned gate.
The objective is to minimize waiting times and queue overflows while maximizing throughput.

\section{State Space}

The state at time $t$ is a tuple $s_t = (t, \mathbf{g}_t, \mathbf{q}_t)$ where:

\begin{itemize}
	\item $t \in \{0, 1, \ldots, T\}$: current time index.
	\item $\mathbf{g}_t \in \{0, 1, \ldots, S_{\max}\}^G$: gate occupancy vector.
	\begin{itemize}
		\item $g_i = 0$: gate $i$ is available.
		\item $g_i = s > 0$: gate $i$ occupied, $s$ time steps remaining.
	\end{itemize}
	\item $\mathbf{q}_t$: queue of waiting flights.
	\begin{itemize}
		\item Each flight $f$ is a tuple $(T_f^{\text{arr}}, r_f, \tau_f, p_f)$, where:
		\begin{itemize}
			\item $T_f^{\text{arr}}$: Actual arrival time.
			\item $r_f \in \{1, \ldots, R\}$: The runway the flight landed on.
			\item $\tau_f \in \mathcal{T}$: Aircraft type from a finite set 
			$\mathcal{T} = \{\text{narrow-body}, \text{wide-body}, 
			\text{regional}, \ldots\}$.
			\item $p_f$: Priority class.
		\end{itemize}
	\end{itemize}
\end{itemize}

\section{Action Space}

At each time step, the decision-maker selects an action $a_t \in \mathcal{A}_t$ where:
%
\begin{equation}
	\mathcal{A}_t = \{(f, g) : f \in \mathbf{q}_t, g \in \text{Available}(\mathbf{g}_t), 
	\text{Compatible}(\tau_f, g)\} \cup \{\text{wait}\}
\end{equation}

An action $(f, g)$ assigns waiting flight $f$ to available gate $g$. The "wait" action defers assignment to the next time step.
\subsection{Aircraft-Gate Compatibility}

Gates are heterogeneous not only in spatial accessibility but also in their 
suitability for different aircraft types.

\paragraph{Hard Constraints (Physical Compatibility):}
Define a binary compatibility matrix $\mathbf{C} \in \{0,1\}^{|\mathcal{T}| \times G}$ 
where:
%
\begin{equation}
	C_{\tau,g} = \begin{cases}
		1 & \text{if aircraft type } \tau \text{ can physically use gate } g \\
		0 & \text{otherwise (structural incompatibility)}
	\end{cases}
\end{equation}
%
The compatibility constraint in the action space enforces 
$\text{Compatible}(\tau_f, g) \equiv C_{\tau_f, g} = 1$.

Examples of structural incompatibilities include:
\begin{itemize}
	\item Wide-body aircraft requiring specific jet bridge configurations
	\item Regional jets unable to use international gates with oversized aprons
	\item Cargo aircraft restricted to remote stands
\end{itemize}

\paragraph{Soft Preferences (Operational Efficiency):}
Beyond physical constraints, certain aircraft-gate pairings are operationally 
preferred. Define a preference matrix $\mathbf{P} \in \mathbb{R}^{|\mathcal{T}| \times G}$ 
where $P_{\tau,g} \geq 0$ represents the operational benefit of assigning type 
$\tau$ to gate $g$.

Preferences may reflect:
\begin{itemize}
	\item Airline hub assignments (same carrier flights clustered for transfers)
	\item Proximity to customs/immigration for international flights
	\item Ground equipment positioning and crew efficiency
\end{itemize}

\section{Transition Dynamics}

The transition from $s_t$ to $s_{t+1}$ involves:

\subsection{Deterministic Time Evolution}

\begin{equation}
	t \to t+1
\end{equation}

\subsection{Gate Occupancy Updates (Heterogeneous Service)}

Gate occupancy is determined by the aircraft's base turnaround time plus a penalty for taxiing distance.
Let $\Delta(r, g)$ be a deterministic distance matrix representing the taxiing time from runway $r$ to gate $g$.

For each gate $i$:
%
\begin{equation}
	g_i^{t+1} = \begin{cases}
		s-1 & \text{if } g_i^t = s > 1 \\
		0 & \text{if } g_i^t = 1 \text{ (service completes)} \\
		\text{BaseService}(\tau_f) + \Delta(r_f, i) & \text{if action assigns 
			flight } f \text{ to gate } i
	\end{cases}
\end{equation}
%
where $\text{BaseService}(\text{type}_f)$ is a random variable sampled from the distribution specific to the aircraft type.

\subsection{Stochastic Schedule-Based Arrivals}

Arrivals are modeled as a perturbed schedule rather than a pure Poisson process.

\noindent Let $\mathcal{T} = \{T_1^{\text{sched}}, T_2^{\text{sched}}, \ldots, T_N^{\text{sched}}\}$ be the master schedule of flight arrival times. The actual arrival time $A_k$ for the $k$-th flight is:
%
\begin{equation}
	A_k = T_k^{\text{sched}} + \delta_k
\end{equation}
%
where $\delta_k$ is a stochastic noise term (e.g., $\delta_k \sim \mathcal{N}(0, \sigma^2)$ or an Erlang delay distribution).

\subsection{Queue Updates}

\begin{equation}
	\mathbf{q}_{t+1} = (\mathbf{q}_t \setminus \{f_{\text{assigned}}\}) \cup \{k : A_k = t+1\}
\end{equation}

\section{Reward Function}

The immediate reward penalizes waiting and encourages efficient assignment:
%
\begin{equation}
	R(s_t, a_t) = -c_{\text{wait}} \cdot |\mathbf{q}_t| 
	- c_{\text{overflow}} \cdot \mathbb{I}[|\mathbf{q}_t| > Q_{\max}] 
	+ c_{\text{assign}} \cdot \mathbb{I}[\text{assignment made}]
	+ \beta \cdot P_{\tau_f, g} \cdot \mathbb{I}[a_t = (f,g)]
\end{equation}
where $\beta > 0$ balances the importance of operational preferences against 
delay minimization. The preference bonus $P_{\tau_f, g}$ is only realized when 
an assignment $(f,g)$ is made, encouraging the agent to match aircraft to 
preferred gates when operationally feasible.

\section{Assumptions}

\begin{assumption}
	Base service times are independent random variables with finite support, 
	dependent on aircraft type $\tau_f$. The distribution parameters vary by 
	type but are known (e.g., narrow-body: $\mathcal{N}(45, 5^2)$ minutes, 
	wide-body: $\mathcal{N}(90, 10^2)$ minutes).
\end{assumption}

\begin{assumption}
	The runway $r_f$ and aircraft type $\tau_f$ for each flight are predetermined 
	by the master schedule and external air traffic control, and are observed 
	upon arrival.
\end{assumption}

\begin{assumption}
	Gates are heterogeneous with respect to both spatial accessibility (taxiing 
	penalties $\Delta(r, g)$) and aircraft compatibility (matrices $\mathbf{C}$, 
	$\mathbf{P}$). The compatibility matrix $\mathbf{C}$ enforces hard physical 
	constraints, while the preference matrix $\mathbf{P}$ represents soft 
	operational efficiencies.
\end{assumption}

\section{Complexity Analysis}

The state space size is:
%
\begin{equation}
	|\mathcal{S}| \approx T \cdot (S_{\max} + \max(\Delta))^G \cdot \binom{F_{\max} + Q_{\max}}{Q_{\max}}
\end{equation}
%
The introduction of the distance matrix $\Delta$ increases the effective support of the service time distribution, but does not increase the dimensionality of the state space, preserving the feasibility of the ADP approach.